---
title: "Lab1"
author: "Brendan Mattina, Ziwei Zhao, Jonathan Moges"
date: "10/26/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
options(tinytex.verbose = TRUE)
install.packages("dplyr")
library("dplyr")
```

## Common Variables

To support our analysis, we defined certain common variables integral to answering the questions posed. Definitions follow below. Throughout the analysis, references to this section will be made when applicable. 

**Voters:** We define voters as people who identified as having definitely voted either in the 2016 general election [turnout16] or in the 2018 primaries [turnout18]. From the data, these are people who chose:
<center>

[1]-Definitely voted in question turnout16,
[1]-Definitely voted in person on Nov 6,
[2]-Definitely voted in person, before Nov 6,
[3]-Definitely voted by mail.,
For both questions, there is an option of "Not completely sure". 

</center>

Respondents who chose this option are asked a new question where they are asked to guess whether they voted, with answers being "Probably did vote" or "Probably did not vote". We decided to be conservative and only analyze responses from people who were certain, rather than also including people who guessed they probably did vote.

## Common Assumptions

Given n = 2500, we can use the Central Limit Theorem to assume the data is Independently and Identically Distributed (IID). 

We also assume the data was randomly drawn from the US population given the sampling methodology provided by the American National Election Studies (ANES). "This was a dual-mode study (face-to-face and Internet), with two independently drawn address-based
samples. The face-to-face component of the study was a complex, stratified, multi-stage cluster sample
of addresses in the 48 contiguous states and Washington DC, while the Internet component was a
simple random sample of eligible addresses in the 50 states and Washington DC." - https://electionstudies.org/wp-content/uploads/2016/02/anes_timeseries_2016_methodology_report.pdf

Voters were defined as respondents who were certain they voted in the 2016 primaries or 2018 general election - detailed methodology in common variables section

Respondents were also asked how not seriously they took answering the questions in the survey - quantified in [nonserious]. They were also asked how honestly they answered questions in the survey [honest]. For simplicity, we chose to not factor the answers to these questions in our analysis. However, further analysis regarding these answers is suggested.

## Question 1: Do US voters have more respect for the police or for journalists?

### Introduce your topic briefly.

Throughout the last four years, few institutions have been more derided than the media and the police. Interestingly, the ANES 2018 Pilot Study included questions that allowed us to define voters and examine their regard for both the media and the police, to begin to answer the question, which institution do American voters respect more.  

Variables that define voters: Please see common variables section.

Variables that Describe Respect for Media and Police: ANES offered participants a thermometer widget to describe their feelings about particular subjects. The participants assign a low temperature to subjects they view unfavorably and a high temperature to subjects they view favorably. The study asked participants “How would you rate the police?” and assigned responses to the variable [ftpolice]. The study also asked participants “How would you rate journalists?” and assigned responses to the variable [ftjounral]. For the purposes of our hypothesis, we believe favorability for journalists is analogous to respect for the media and favorability for the police is analogous to respect for the police. We compared [ftjournal] and [ftpolice] for each voter to test our hypothesis. 

### Perform an exploratory data analysis (EDA) of the relevant variables.

```{r problem 1}

data <- read.csv('anes_pilot_2018.csv')

#tables sum to 2500, or length of the sample and don't include any surprises
table(data$turnout18)
table(data$turnout16)

#extract voter sample
vsamp1 = data[data$turnout18 < 4 | data$turnout16 == 1,]

#now EDA check to see that lengths are the same
nrow(vsamp1)
length(vsamp1$ftpolice)

#take a look at variable summaries 
summary(vsamp1$ftpolice)
summary(vsamp1$ftjournal)

#see negative numbers so will have to clear that from the sample
vsamp1 <- vsamp1[vsamp1$ftjournal >= 0,]

#new sample size and summary
n = nrow(vsamp1)
summary(vsamp1$ftjournal)

#histograms
hist(vsamp1$ftpolice, main = "Histogram of FTPOLICE Variable", 
     xlab = 'Police Favoritability Scores (0-100)')
hist(vsamp1$ftjournal, main = "Histogram of FTJOURNAL Variable", 
     xlab = 'Journalist Favoritability Scores (0-100)')
```

### Based on your EDA, select an appropriate hypothesis test.

The data, in this case, is ordinal as the study directs participants to use a temperature scale to assess favorability of the police and journalists. In this case, the temperature scale is relative to each participant. Ultimately, we only know that some temperatures are higher than others, we have do not know what each temperature means. 

Additionally, we know the data was randomly drawn from the US population (more in our common assumptions section). Since we are comparing how each participant scores both [ftjournal] and [ftpolice], the data is not independent. 

Considering ordinal and non-independent, paired data, we settled on using a Sign Test to asses our hypothesis, with $H_o = .5$ and $H_a \neq .5$. In a one-sample sign test, the test statistics follow a binomial distribution. This test has less statistical power than the Wilcox signed-rank test, but requires fewer assumptions (random, paired sample). Histograms visually suggested only a slight preference for police over media, so with no other convincing support for an alternate test, we used a two-sided test.  

### Conduct your test. Explain your results. 

```{r problem 1 hypothesis test}

#conduct rank sign test to determine significance
rd = sum(vsamp1$ftpolice > vsamp1$ftjournal, na.rm=T)
trials1 = sum(vsamp1$ftpolice > vsamp1$ftjournal | vsamp1$ftpolice < vsamp1$ftjournal)
binom.test(rd , trials1)

#effect size - common language: 
  #data_for/total_data. total_data = people w/ an expressed difference

es_CommLang1 = rd/trials1 #.56 - relatively low
es_CommLang1
#effect size - simple correlation: r = (data in favor - data against)/(total data)

es_Corr1 = (rd-(trials1-rd))/trials1 # .13 - low correlation
es_Corr1
```

We conducted a two-sided Sign Test with $H_o = .5$, or, when there was a difference, half of voters hold a more favorable view of the police and half of voters hold a more favorable view of the media, with $H_a \neq .5$. We set our confidence level at .95. The test returned $p = 7.12 x 10^{-9}$, meaning that we rejected $H_o$. 

This result, combined with the observed practical significance, the common language effect size, of .56, suggest that voters may have slightly more respect for the police, although correlation of .13 is low. Ultimately more study is needed to verify voters preference for the police over the media.    

```{r}
AA = read.csv("anes_pilot_2018.csv")
```
## Question 2: Are Republican voters older or younger than Democratic voters?

### Introduce your topic briefly.  (5 points)
Explain how your variables are operationalized.  Comment on any gaps that you can identify between your operational definitions and the concepts you are trying to study.

We are interested in understanding if Republican voters are older or younger than Democratic voters. To understand this, we must first define certain concepts key to this question. 

1) Older or younger. In the survey, respondents are asked "In what year were you born?" through variable [birthyr]. Birth year allows respondents to choose their birth year with values ranging from 1927 to 2000. 

2) Democratic or Republican. Respondents are asked if they "usually think of themselves as a Democrat, a Republican, an independent, or what?" through two questions - [pid1d] and [pid1r]. We observed that [pid1r] asks the same question as [pid1d] but randomly assigns the placement Democrat and Republican in positions 1 and 2. We interpreted these results as the respondents' party association. 

3) Voters. Refer to common variables section for definition and treatment. 

```{r}
#create dataset of voters who think of themselves as...
#...Democratic or Republican and who voted in 2016 or 2018
BB = AA[(AA$pid1d==1 | AA$pid1r==2) | (AA$pid1d==2 | AA$pid1r==1)
        & (AA$turnout16==1|(AA$turnout18>=1&AA$turnout18<=3)),]
```

### Perform an exploratory data analysis (EDA) of the relevant variables. (5 points)
This should include a treatment of non-response and other special codes, basic sanity checks, and a justification for any values that are removed.  Use visual tools to assess the relationship among your variables and comment on any features you find.

```{r}

#create "dems" to represent Democratic voters and 
#"reps" to represent Republican voters who voted in 2016 or 2018. 
BB$dems <- (BB$pid1d == 1 | BB$pid1r == 2)
BB$reps <- (BB$pid1d == 2 | BB$pid1r == 1)

#create "demrep crosstab to represent dems and reps. 
#No unexpected values or nulls in raw data. Create "demrep" crosstab
demrep <- table(BB$dems, BB$reps)

#remove zeros
demrep[demrep == 0] <- NA

#rename columns and rows for understanding
colnames(demrep) = c("Republicans", "Democrats")
rownames(demrep) = c("Democrats", "Republicans")
```

```{r}

#validating if "dems", "reps" and "demrep" work as designed
table(BB$pid1d, BB$pid1r, BB$dems, BB$reps)
```

```{r}

#no unexpected values or nulls in birthyr
table(BB$birthyr)
```

```{r}

#data is slightly bimodally distributed
hist(BB$birthyr, main = "Distribution of Birth Years of Respondents", 
     xlab = "Birth Year", ylab = "Number of Respondents") 
```

From the cross tab and histogram, we observe that the birth year data is slightly bimodally distributed and 52.4% of respondents are Republican while 47.6% are Democrats. No outliers or null data were observed. Prior to focusing only on Democratic and Republican respondents, the pid1d and pid1r data had outliers, but were removed during the EDA, resulting in normally distributed data. 

```{r}

#conduct variance test to see if variances are equal
var.test(BB$birthyr, demrep)
```

### Based on your EDA, select an appropriate hypothesis test.  (5 points)
Explain why your test is the most appropriate choice.  List and evaluate all assumptions for your test.

Additionally, our data is unpaired and parametric. We chose an unpaired t-test. Given n = 1364 when focusing on Democrats and Republican respondents, we can use the CLT to assume normality. 

By conducting a variance test, we compare the variances of the desired groups of the t-test: birthyr and demrep. We observe the p-value is less than the significance level of .05. This means we'll assume unequal variances in our t-test and conduct a two-tailed test

### Conduct your test. (5 points)
Explain (1) the statistical significance of your result, and (2) the practical significance of your result.  Make sure you relate your findings to the original research question.

```{r}

#conduct unpaired two-sided t-test
t.test(BB$birthyr, demrep, alternative = "two.sided", paired = FALSE, 
       var.equal = FALSE, conf.level = .95)
```

We conducted a t-test that computed a p-value <.05 where the null hypothesis is that Republican voters would be older than Democratic voters and the alternative hypothesis is that Republican voters would be younger than Democratic voters.

```{r}

#compute the standard deviation of birth year
sd(BB$birthyr)

#create variables "birthyr_2" and "demrep_2" to compute effect size and correlation
table(BB$birthyr)
BB$birthyr_2 <- sum(BB$birthyr)
demrep_2 <- sum(BB$pid1d & BB$pid1r)

#cohen's D: (1965.489-682)/sqrt(16.55265) = .708 - somewhat large effect size
effsize <- (BB$birthyr_2/demrep_2) 

#(682-(1965.489-682))/1965.489 =-.3 - slightly negatively correlated. Weak linear relationship
cor <- (demrep_2-(BB$birthyr_2-demrep_2))/BB$birthyr_2 
```

Given the computed effect size of 1.44, there is a very large effect and significant correlation.

```{r}
A = read.csv("anes_pilot_2018.csv")
```

## Question 3: Do a majority of independent voters believe that the federal investigations of Russian election interference are baseless?

### Introduce your topic briefly.  (5 points)

We are interested in the opinion of independent voters' opinions of the federal investigations, specifically if the majority of them think of the investigation as baseless. There are several concepts that need to be defined:

1) Independent. In the survey, respondents are asked whether they think of themselves as closer to being a Republican, Democrat, or independent through one of two questions - pid1d or pid1r. These two questions have the same four answers, Democrat-[1], Republican-[2], independent-[3], or something else-[4], except in different orders. Each respondent only answers one of the two questions. Therefore, we defined independent voters as the population of respondents who either chose [3] for pid1d or chose [3] for pid1r.

2) Think of the federal investigations of Russian election interference as baseless. There are three variables we found that pertain to this question:
    1. russia16 - [1]-Russia probably did interfere in the 2016 presidential election to try to help Donald Trump win, or [2]-This probably did not happen
    2. muellerinv - Do you approve, disapprove, or neither approve nor disapprove of Robert Mueller’s investigation of Russian interference in the 2016 election?
    3. coord16 - [1]-Donald Trump’s 2016 campaign probably coordinated with the Russians, or [2]-This probably did not happen
While all three are relevant, we interpret our research question as more about whether the interference itself happened, rather than the quality of the investigation. It is possible for someone to believe the investigation to be baseless, yet approve of the investigation, because they want the investigation to disprove rumors of interference; or someone to to believe that there probably was interference, but do not approve of the investigation because they disprove of the methodology of the investigation, or Robert Mueller as a person. Thus, we only use variables russia16 and coord16, and not muellerinv. We define someone believing the investigation to be not baseless to be someone who either chose [1] for russia16, or [1] for coord16. This means that in our definition, someone believes the investigation to be baseless if they chose [2] for both russia16 and coord16. We create a binomial new variable "baseless" that takes values 0 or 1.
  

### Perform an exploratory data analysis (EDA) of the relevant variables. (5 points)

```{r}

# confirming that only one of these two questions about party preference is answered
table(A$pid1d, A$pid1r) 

# no unexpected values in turnout 16 and turnout18
table(A$turnout16, A$turnout18) 

# make subset data of only independent voters who voted in 2016 OR 2018
B = A[(A$pid1r==3 | A$pid1d==3) & (A$turnout16==1 | (A$turnout18>=1 & A$turnout18 <=3)),]

# make new "baseless" variable for people who chose 2 for both russia16 and coord16
B$baseless <- (B$russia16==2 & B$coord16==2)

# Checking baseless variable against russia16 and coord16 to confirm it was made as we desired
table(B$russia16, B$coord16, B$baseless)

# Count number of independent voters believing the investigation to be baseless
baseless <- sum(B$baseless)

# Count total independent voters
pop <- nrow(B)

# Plot the distribution
plot(0:nrow(B), dbinom(0:nrow(B), size=nrow(B), prob=baseless/pop),
     type='h',
     main='Distribution of Variable Baseless',
     ylab='Probability',
     xlab ='# Independent voters believing baseless investigation',
     lwd=3)
```

## Question 4: Was anger or fear more effective at driving increases in voter turnout from 2016 to 2018?

### Introduce your topic briefly. 
This feels like an intensely emotional time for American politics. The prevailing emotions of anger and fear seem to be driving Americans towards increased, albeit erratic, political activity. But which emotion drives more new voters to the polls? 

**Variables that define new voters:** In addition to the variables discussed in our common variables section, we leveraged the [turnout16b] variable to extract any and all voters voted in the 2018 Midterm but not the 2016 General Elections. If a participant was unsure about whether they voted in 2016, they were asked if they probably voted or probably did not vote. If the participant answered that they definitely did not vote in 2016, [turnout16] = 2, probably did not vote, [turnout16] = 3 & [turnout16B] = 2, but definitely voted in 2018, [turnout18] < 4, then we included them in our sample of 'new voters'. Finally we used new voters' birth year, as coded by [birthyr] (for a discussion on this variable please see question 2) to ensure that new voters were at least 18 in 2016 (born in 1998 or before) which means they legally had the chance to vote and did not. We understand that filtering on $[birthyr] \leq 1998$ may allow for the addition of voters who were born after November 6, 1998 and were therefore unable to vote in 2016. However, we're comfortable assuming that risk for the benefit of including the remaining 85% of birth year 1998 in our analysis.     

**Variables that describe anger or fear:** ANES presented participants with a 'global emotional battery' that asked participants about how they 'felt about the country these days', in terms of various emotions. ANES used a likert scale, 1-"Not at all" to 5-"Extremely" to capture the intensity of each participant's emotion. We found that the variables [geangry] and [geafraid] most succinctly captured participants level of anger and fear. Given the ordinal nature of the data, we unfortunately could not integrate participant anger or fear as captured in other emotion batteries included in the survey; however, these additional variables should be included in subsequent studies. 

### Perform an exploratory data analysis (EDA) of the relevant variables. (5 points)
```{r, problem 4 start}

data <- read.csv('anes_pilot_2018.csv')

table(data$turnout16 > 1 | (data$turnout16 == 3 & data$turnout16b == 2), data$turnout18 < 4)

```

```{r, load data and EDA}

#identify voters who definitely or probvably did not vote in the General Election.
vsamp = data[(data$turnout16 > 1 | (data$turnout16 == 3 & data$turnout16b == 2)) 
             & data$turnout18 < 4 & data$birthyr <= 1998,]

#summary of the anger/fear variables, notice negative values. 
summary(vsamp$geangry)
summary(vsamp$geafraid)

#now remove non-answer values from each category
vsamp = vsamp[ vsamp$geangry != -7 & 
               vsamp$geafraid != -7,] 

#double check integrity of vsamp
length(vsamp$geafraid) 
length(vsamp$geangry) 

table(vsamp$geafraid,vsamp$geangry)

#data histograms

hist(vsamp$geangry, main = "Histogram of GEANGRY Variable", 
     xlab = 'Global Emotion - Anger Scores (1-5)')
hist(vsamp$geafraid, main = "Histogram of GEAFRAID Variable", 
     xlab = 'Global Emotion - Fear Scores (1-5)')

```

### Based on your EDA, select an appropriate hypothesis test.  (5 points)

The data in this case is ordinal as participants used a likert scale to describe the intensity of their emotion. The scale values are relative to the participant and therefore have no stand-alone value.

Additionally, we know that the data was randomly drawn from the US population (more in our common assumptions section). Since we are comparing how each participant scores both [geangry] and [geafraid], the data is not independent. 

Considering ordinal and non-independent, paired, data, we decided to use a Sign Test to asses our hypothesis, with $H_o = .5$ and $H_a \neq .5$. In an one-sample sign test, the test statistics follow a binomial distribution. This test has less statistical power than the Wilcox signed-rank test, but requires fewer assumptions (random, paired sample) the histograms suggested no obvious difference in the median of each variable, so with no other convincing support for an alternate test, we used a two-sided test.

### Conduct your test. Explain your results.
```{r}

#now conduct signed test that anger was a greater motivator than fear
rel_diff = sum(vsamp$geangry > vsamp$geafraid, na.rm=T)
table(rel_diff)

#ttoal trials, essentially any time a particiapnt indicated...
#...a difference between their anger and their fear
trials = sum(vsamp$geangry > vsamp$geafraid | vsamp$geangry < vsamp$geafraid)
binom.test(rel_diff , trials)

#effect size - common language: data_for/total_data. total_data = people w/ an expressed difference

es_CommLang = 24/46 #.52, slightly favors anger over fear 
es_CommLang

#effect size - simple correlation: r = (data in favor - data against)/(total data)

es_Corr = (24-(46-24))/46 # 0.0435 - minimal correlation
es_Corr

```

We conducted a two-sided Sign Test with $H_o = .5$, or, when there was a difference, half of new voters expressed more anger and half of new voters expressed more fear, with $H_a \neq .5$. We set our confidence level at .95. The test returned $p = .883$, meaning that we failed to reject $H_o$. 

This result, combined with the observed practical significance, the common language effect size, of .52 (slightly favoring anger over fear), and general correlation of 0.0435 (very low correlation again favoring anger), suggest that practically speaking, new voters are as angry as they are afraid and neither emotion had a greater impact at driving new voter turnout in 2018. More study is needed to identify emotions that motivates political activity. 

## Question 5: Select a fifth question that you believe is important for understanding the behavior of voters

### Clearly argue for the relevance of this question.  (10 points)

Our research question is: are female voters more liberal than male voters? Firstly, gender is one of the most common ways to segment the population in politics. Also, gender issues such as abortion have been historically political for which the two parties take polarizing stances. Therefore, we find it worthwhile to investigate whether there is statistical difference in ideology in men and women. 

For this question, we are using two variables:

1) gender - we will use the profile variable which takes value [1] for male, and [2] for female. A potential gap is that the question does not specify whether the respondent should answer with their biological sex or the gender with which they identify. There is also no third option for people who may not wish to disclose their gender. 

2) lcself - this variable records respondents' answer to the question of how liberal or conservative they identify. There are seven categories, from [1]-very liberal to [7]-very conservative, with [4] being neither liberal nor conservative. This is a very subjective question, as all questions asking about ideology would be. Each person has their own definition of liberal or conservative, and it is also possible for people to not feel comfortable expressing extreme views, thus choosing something more moderate than how they actually feel. But we find this be the most appropriate variable to help us answer our question.

To keep voters, see common assumption regarding treatment of voters.

### Perform EDA and select your hypothesis test (5 points)

Perform an exploratory data analysis (EDA) of the relevant variables.

This should include a treatment of non-response and other special codes, basic sanity checks, and a justification for any values that are removed.  Use visual tools to assess the relationship among your variables and comment on any features you find.

Based on your EDA, select an appropriate hypothesis test.
Explain why your test is the most appropriate choice.  List and evaluate all assumptions for your test.

```{r}

# Cross tabulate gender and lcself variable
table(A$gender, A$lcself)

# Make subset data for voters who answered the ideology question
C = A[(A$lcself >=1 & A$lcself <= 7) & (A$turnout16==1 | (A$turnout18>=1 & A$turnout18 <=3)),]

# Cross tabulate gender and lcself variable again using the subset data
table <- table(C$gender, C$lcself)
table
colnames(table) = c("Very liberal", "Somewhat liberal", 
                    "Closer to liberal", "Neither liberal nor conservative", 
                    "Closer to conservative", "Somewhat conservative", "Very conservative")
rownames(table)=c("Male", "Female")
table2 <- prop.table(table, 1)
table2

# Make plot
par(mar=c(8, 4.1, 4.1, 2.1))
barplot(table2,
        main = "Tabulation of Ideology by Gender by %",
        ylim=c(0,0.25),
        beside=TRUE,
        xlab="",
        las=2,
        col=c("skyblue","maroon"),
        cex.names=0.6)
legend("topright",
       title="Gender",
       legend=c("Male", "Female"),
       fill=c("skyblue", "maroon"),
       box.lty=0, cex=0.8)
```

From the bar plot, we can see that there are more females than males in four categories from very liberal to neither liberal nor conservative, as well as a few more females in closer to conservative. There are more males choosing somewhat conservative and very conservative. 
Because our data is unpaired and non-parametric, we choose to conduct a Wilcoxon Rank-Sum Test that requires two assumptions:
1) Ordinal scale. Our data is categorical that takes seven values. It is ordinal. 
2) IID - See common assumptions.
### Conduct your test.
```{r}
wilcox.test(C$lcself~C$gender)
cor.test(C$lcself, C$gender, method='spearman')
```
The P-value is very small, thus we reject the null hypothesis that it's equally likely for females to be more liberal than for males to be more liberal on the scale of 1 to 7.
The spearman correlation is -0.12. The negative sign indicates that as the dependent variable increases (from [1]-male to [2]-female), the value of lcself decreases, or closer to liberal. The correlation is not very big.
### Conclusion
Through conducting a Wilcoxon Rank-Sum Test, our conclusion is that it is not equally likely for female voters to be more liberal than for male voters to be liberal, and the results of the Spearman Correlation suggest that women tend to be more liberal. This test is based on the voters' self assessment of how liberal or conservative they are, so more accurately, we can say the correlation suggests that female voters are more likely to identify themselves as liberal than men. Ultimately more testing is needed to confirm this suggestion.